### 2020-01-12 grand_challenge/astro_grand_challenge.tex
change size
final version

\usepackage{hyperref}
\usepackage{mathpazo}

@@

\begin{center}
{\Large Challenges in Modeling Astrophysical Thermonuclear Explosions} \\[0.25em]
Michael Zingale (Stony Brook)
\end{center}

The end states of stellar evolution result in exotic compact objects:
white dwarfs, neutron stars, and black holes.  When these remnants are
in binary systems, interactions with the companion star can lead to a
variety of different thermonuclear explosions: classical novae, X-ray
bursts (XRBs), and thermonuclear (Type Ia) supernovae (SN Ia).  SN Ia
and novae are important sites of nucleosynthesis in the galaxy; XRBs
allow us to probe the physics of dense nuclear matter; and SN Ia are
important distant indicators for measuring the expansion history of
the Universe.  Simulations on these events have been ongoing for
decades and allow us to understand the physics of the explosions,
interpret the observations, and test different progenitor models for
viability.  Next generation supercomputers will allow us to push these
simulations to unprecedented realism.

Modeling these events requires multiscale and multiphysics simulations.
Spatial scales range from the size of the compact object
down to the burning or dissipation scale.  Temporary scales range from
the millions of years (SN Ia) to hours (XRBs) of evolution that
brought the star to the brink of explosions to the seconds-long
explosion.  Models must incorporate hydrodynamics, nuclear reactions,
gravity, thermal conduction, magnetic fields, and radiation.

While a variety of approximations are used in astrophysics to model
hydrodynamic flows, for stellar explosions, conservative finite-volume
methods are the most popular.  Adaptive mesh refinement (AMR) allows
us to increase the range of lengthscales captures.  Our group has been
developing the AMReX Astrophysics Suite of open-source AMR
multiphysics hydrodynamics
codes\footnote{\url{https://github.com/AMReX-Astro/}} for over a
decade.  Two codes in particular address stellar explosions: MAESTROeX
is a low Mach number hydrodynamics code for modeling the highly
subsonic stellar convection leading up to explosion and Castro is a
fully compressible radiation hydrodynamics code that efficiently
models the explosive phase.  These codes share a common framework
(adaptive mesh library, microphysics routines), are written in C++ and Fortran,
and use MPI+OpenMP or MPI+CUDA.

@@

To highlight the way that new computer architectures enable new
science, we focus on our current work on XRBs.  Our interest is in
modeling the nucleosynthesis as a burning front propagates laterally
through the accreted fuel layer on a neutron star.  Due to the range
of length and timescales, various approximations have been used in the
literature.  Our group started performing two-dimensional
hydrodynamical simulations of the flame propagating using the NERSC
Cori and OLCF Titan (on CPUs)  machines over a year ago.  These calculations were
the first to perform full hydrodynamics in both the lateral and
vertical directions, allowing us to understand the complex dynamics of
the flame front with realistic microphysics.  The simulations were
challenging, and we needed to make approximations to both the length
scales (using artificially high neutron star rotation rates to compress
the lateral scales) and time scales (boosting the flame speed).
Nevertheless, we were able to complete the first resolved calculations
of hydrodynamical flames in X-ray bursts.  Taking advantage of GPUs,
we are now able to eliminate these approximations in our current runs on the
OLCF Summit machine.  The same code is used,
but running entirely on the GPUs, it is 5--10$\times$ faster than
CPUs.  We are on target to perform the first three-dimensional
simulations this year as well as to explore the more interesting case
of mixed H/He bursts, which involve more complicated nuclear
reactions.

>>>>

\usepackage{hyperref}
\usepackage{times}

\usepackage[para]{footmisc}

@@

\begin{center}
{\large Challenges in Modeling Astrophysical Thermonuclear Explosions} \\[0.25em]
Michael Zingale (Stony Brook)
\end{center}

The end states of stellar evolution result in exotic compact objects:
white dwarfs, neutron stars, and black holes.  When these remnants are
in binary systems, interactions with the companion star can lead to a
variety of different thermonuclear explosions: classical novae, X-ray
bursts (XRBs), and thermonuclear (Type Ia) supernovae (SN Ia).  SN Ia
and novae are important sites of nucleosynthesis in the galaxy; XRBs
allow us to probe the physics of dense nuclear matter; and SN Ia are
important distance indicators for measuring the expansion history of
the Universe.  Simulations on these events have been ongoing for
decades and allow us to understand the physics of the explosions,
interpret the observations, and test different progenitor models for
viability.  Next generation supercomputers will allow us to push these
simulations to unprecedented realism.

Modeling these events requires multiscale and multiphysics simulations.
Spatial scales range from the size of the compact object
down to the burning or dissipation scale.  Temporal scales range from
the millions of years (SN Ia) to hours (XRBs) of evolution that
brought the star to the brink of explosion to the seconds-long
explosion.  Models must incorporate hydrodynamics, nuclear reactions,
gravity, thermal conduction, magnetic fields, and radiation.

While a variety of approximations are used in astrophysics to model
hydrodynamic flows, for stellar explosions, conservative finite-volume
methods are the most popular.  Adaptive mesh refinement (AMR) allows
us to increase the range of length scales captured.  Our group develops
the AMReX Astrophysics Suite of open-source AMR
multiphysics hydrodynamics
codes\footnote{\url{https://github.com/AMReX-Astro/}}.  Two codes in particular address stellar explosions: MAESTROeX
is a low Mach number hydrodynamics code for modeling the highly
subsonic stellar convection leading up to explosion and Castro is a
fully compressible radiation hydrodynamics code that efficiently
models the explosive phase.  These codes share a common framework
(AMR library, microphysics routines), are written in C++ and Fortran,
and use MPI+OpenMP or MPI+CUDA.

@@

To highlight the way that new computer architectures enable new
science, we focus on our current work on XRBs.  Our interest is in
modeling the nucleosynthesis as a burning front propagates laterally
through the accreted fuel layer on a neutron star.  Due to the range
of length and time scales, various approximations have been used in the
literature.  Our group started performing 2D
hydrodynamical simulations of the flame propagating using the NERSC
Cori and OLCF Titan (on CPUs)  machines over a year ago.  These calculations were
the first to perform full hydrodynamics in both the lateral and
vertical directions, allowing us to understand the complex dynamics of
the flame front with realistic microphysics.  The simulations were
challenging, and we needed to make approximations to both the length
scales (using artificially high neutron star rotation rates to compress
the lateral scales) and time scales (boosting the flame speed).
  Taking advantage of GPUs,
we are now able to eliminate these approximations in our current runs on the
OLCF Summit machine.  The same code is used,
but running entirely on the GPUs, it is 5--10$\times$ faster than
CPUs.  We are on target to perform the first three-dimensional
simulations this year as well as to explore the more interesting case
of mixed H/He bursts, which involve more complicated nuclear
reactions.
